---
title: "Analyze structural shift"
author: "Tyler Starr"
date: "1/26/2022"
output:
  github_document:
    toc: true
    html_preview: false
editor_options: 
  chunk_output_type: inline
---
This notebook analyzes structural shifts in the ACE2-bound RBD structure, and compares to functional shifts in mutational effects.

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("yaml","data.table","tidyverse","gridExtra","egg","bio3d","ggrepel")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("config.yaml")

#make output directory
if(!file.exists(config$structural_shifts_dir)){
  dir.create(file.path(config$structural_shifts_dir))
}

```
Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

Define colorblind-friendly palette
```{r define_color_palette}
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", 
               "#0072B2", "#D55E00", "#CC79A7")
# The palette with black
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                "#0072B2", "#D55E00", "#CC79A7")
```

## Setup

Read in PDBs. These were aligned to minimize RBD Calpha RMSD using PyMol. For beta, omicron structures, this was performed for both RBD units in the asymmetric unit. (These structures also included mAbs used in the crystallization.) RBD and ACE2s were then split into separate objects/pdbs. The pdbs in these aligned coordinates were then output. Structures include WH1 (6m0j), N501Y (7ekf), beta (beta1=7ekg, beta2,3 not yet published), and omicron (omicron 1,2 not yet published, omicron3=7wbp).

```{r input_data}
WH1 <- read.pdb(file=config$pdb_WH1)
N501Y <- read.pdb(file=config$pdb_N501Y)
beta1 <- read.pdb(file=config$pdb_beta1)
beta2 <- read.pdb(file=config$pdb_beta2)
beta3 <- read.pdb(file=config$pdb_beta3)
delta1 <- read.pdb(file=config$pdb_delta1)
delta2 <- read.pdb(file=config$pdb_delta2)
omicron1 <- read.pdb(file=config$pdb_omicron1)
omicron2 <- read.pdb(file=config$pdb_omicron2)
omicron3 <- read.pdb(file=config$pdb_omicron3)
```

## Calculate pairwise C-alpha, beta, and gamma distances for each structure compared to WH1

Note, the Cgamma calc excludes residues with no gamma-atom (alanine and glycine) or beta-branching (and thus a gamma1 and gamma2 atom: isoleucine, threonine, valine)

```{r calc_Ca_dists}
RBD_sites <- data.table(read.csv(file=config$RBD_sites))

for(voc in c("N501Y","beta1","beta2","beta3","delta1","delta2","omicron1","omicron2","omicron3")){
  RBD_sites[,paste(voc,"_v_WH1_Ca_dist",sep=""):=as.numeric(NA)]
  for(i in 1:nrow(RBD_sites)){
    site <- RBD_sites[i,site]
    WH1_calpha <- WH1$atom[WH1$atom$resno==site & WH1$atom$elety=="CA",]
    voc_calpha <- get(voc)$atom[get(voc)$atom$resno==site & get(voc)$atom$elety=="CA",]
    if(nrow(WH1_calpha)==1 & nrow(voc_calpha)==1){
      RBD_sites[i,paste(voc,"_v_WH1_Ca_dist",sep="") := sqrt((WH1_calpha$x - voc_calpha$x)^2 + (WH1_calpha$y - voc_calpha$y)^2 + (WH1_calpha$z - voc_calpha$z)^2)] 
    }else{
      #print(paste("no distance for", voc, "site",site))
    }
  }
}

```

```{r calc_Cb_dists}
for(voc in c("N501Y","beta1","beta2","beta3","delta1","delta2","omicron1","omicron2","omicron3")){
  RBD_sites[,paste(voc,"_v_WH1_Cb_dist",sep=""):=as.numeric(NA)]
  for(i in 1:nrow(RBD_sites)){
    site <- RBD_sites[i,site]
    WH1_cbeta <- WH1$atom[WH1$atom$resno==site & WH1$atom$elety=="CB",]
    voc_cbeta <- get(voc)$atom[get(voc)$atom$resno==site & get(voc)$atom$elety=="CB",]
    if(nrow(WH1_cbeta)==1 & nrow(voc_cbeta)==1){
      RBD_sites[i,paste(voc,"_v_WH1_Cb_dist",sep="") := sqrt((WH1_cbeta$x - voc_cbeta$x)^2 + (WH1_cbeta$y - voc_cbeta$y)^2 + (WH1_cbeta$z - voc_cbeta$z)^2)] 
    }else{
      #print(paste("no distance for", voc, "site",site))
    }
  }
}

```

```{r calc_Cg_dists}
for(voc in c("N501Y","beta1","beta2","beta3","delta1","delta2","omicron1","omicron2","omicron3")){
  RBD_sites[,paste(voc,"_v_WH1_Cg_dist",sep=""):=as.numeric(NA)]
  for(i in 1:nrow(RBD_sites)){
    site <- RBD_sites[i,site]
    WH1_cgamma <- WH1$atom[WH1$atom$resno==site & WH1$atom$elety %in% c("CG","SG","OG"),]
    voc_cgamma <- get(voc)$atom[get(voc)$atom$resno==site & get(voc)$atom$elety %in% c("CG","SG","OG"),]
    if(nrow(WH1_cgamma)==1 & nrow(voc_cgamma)==1){
      RBD_sites[i,paste(voc,"_v_WH1_Cg_dist",sep="") := sqrt((WH1_cgamma$x - voc_cgamma$x)^2 + (WH1_cgamma$y - voc_cgamma$y)^2 + (WH1_cgamma$z - voc_cgamma$z)^2)] 
    }else{
      #print(paste("no distance for", voc, "site",site))
    }
  }
}

```

Average across all atoms in each residue

```{r calc_all-atom_dists}
for(voc in c("N501Y","beta1","beta2","beta3","delta1","delta2","omicron1","omicron2","omicron3")){
  RBD_sites[,paste(voc,"_v_WH1_all_atom_dist",sep=""):=as.numeric(NA)]
  for(i in 1:nrow(RBD_sites)){
    site <- RBD_sites[i,site]
    WH1_atoms <- WH1$atom[WH1$atom$resno==site,]
    voc_atoms <- get(voc)$atom[get(voc)$atom$resno==site,]
    if(nrow(WH1_atoms)>1 & nrow(voc_atoms)>1 & nrow(WH1_atoms)==nrow(voc_atoms)){
      if(WH1_atoms$resid[1]==voc_atoms$resid[1]){
        RBD_sites[i,paste(voc,"_v_WH1_all_atom_dist",sep="") := mean(sqrt((WH1_atoms$x - voc_atoms$x)^2 + (WH1_atoms$y - voc_atoms$y)^2 + (WH1_atoms$z - voc_atoms$z)^2))] 
      }
    }
  }
}

```

Line plots of Calpha, beta, gamma, atom-averaged displacement compared to WH1 structure

```{r line_plots_Calpha-dist_v_WH1, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
calpha <- melt(RBD_sites[,.(site,N501Y_v_WH1_Ca_dist,beta1_v_WH1_Ca_dist,beta2_v_WH1_Ca_dist,beta3_v_WH1_Ca_dist,delta1_v_WH1_Ca_dist,delta2_v_WH1_Ca_dist)], id.vars=c("site"))
calpha[,rep:=1]
calpha[variable=="N501Y_v_WH1_Ca_dist",variable:="N501Y"]
calpha[variable=="beta1_v_WH1_Ca_dist",variable:="Beta"]
calpha[variable=="beta2_v_WH1_Ca_dist",rep:=2]
calpha[variable=="beta2_v_WH1_Ca_dist",variable:="Beta"]
calpha[variable=="beta3_v_WH1_Ca_dist",rep:=3]
calpha[variable=="beta3_v_WH1_Ca_dist",variable:="Beta"]
calpha[variable=="delta1_v_WH1_Ca_dist",variable:="Delta"]
calpha[variable=="delta2_v_WH1_Ca_dist",rep:=2]
calpha[variable=="delta2_v_WH1_Ca_dist",variable:="Delta"]

#define colors for each bg
group.colors <- c("Wuhan-Hu-1" = cbPalette[1], "N501Y" = cbPalette[3], "E484K" = cbPalette[5], "Beta"=cbPalette[6], "Delta"=cbPalette[7], "Omicron"=cbPalette[2])

#define order for plotting of bgs
calpha$variable <- factor(calpha$variable,levels=c("E484K","N501Y","Beta","Delta"))

ggplot(data=calpha[rep==1 & site %in% seq(334,515),], aes(x=site, y=value, color=variable))+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("C-alpha distance versus Wuhan-Hu-1 [6m0j] (A)")+
  geom_text_repel(aes(label=ifelse(((value > 0.75) & variable=="Beta"),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/Calpha_v_WH1.pdf",sep=""),useDingbats=F))
```

```{r line_plots_Cbeta-dist_v_WH1, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
cbeta <- melt(RBD_sites[,.(site,N501Y_v_WH1_Cb_dist,beta1_v_WH1_Cb_dist,beta2_v_WH1_Cb_dist,beta3_v_WH1_Cb_dist,delta1_v_WH1_Cb_dist,delta2_v_WH1_Cb_dist)], id.vars=c("site"))
cbeta[,rep:=1]
cbeta[variable=="N501Y_v_WH1_Cb_dist",variable:="N501Y"]
cbeta[variable=="beta1_v_WH1_Cb_dist",variable:="Beta"]
cbeta[variable=="beta2_v_WH1_Cb_dist",rep:=2]
cbeta[variable=="beta2_v_WH1_Cb_dist",variable:="Beta"]
cbeta[variable=="beta3_v_WH1_Cb_dist",rep:=3]
cbeta[variable=="beta3_v_WH1_Cb_dist",variable:="Beta"]
cbeta[variable=="delta1_v_WH1_Cb_dist",variable:="Delta"]
cbeta[variable=="delta2_v_WH1_Cb_dist",rep:=2]
cbeta[variable=="delta2_v_WH1_Cb_dist",variable:="Delta"]

#define order for plotting of bgs
cbeta$variable <- factor(cbeta$variable,levels=c("E484K","N501Y","Beta","Delta"))

ggplot(data=cbeta[rep==1 & site %in% seq(334,515),], aes(x=site, y=value, color=variable))+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("C-beta distance versus Wuhan-Hu-1 [6m0j] (A)")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/Cbeta_v_WH1.pdf",sep=""),useDingbats=F))
```


```{r line_plots_Cgamma-dist_v_WH1, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
cgamma <- melt(RBD_sites[,.(site,N501Y_v_WH1_Cg_dist,beta1_v_WH1_Cg_dist,beta2_v_WH1_Cg_dist,beta3_v_WH1_Cg_dist,delta1_v_WH1_Cg_dist,delta2_v_WH1_Cg_dist)], id.vars=c("site"))
cgamma[,rep:=1]
cgamma[variable=="N501Y_v_WH1_Cg_dist",variable:="N501Y"]
cgamma[variable=="beta1_v_WH1_Cg_dist",variable:="Beta"]
cgamma[variable=="beta2_v_WH1_Cg_dist",rep:=2]
cgamma[variable=="beta2_v_WH1_Cg_dist",variable:="Beta"]
cgamma[variable=="beta3_v_WH1_Cg_dist",rep:=3]
cgamma[variable=="beta3_v_WH1_Cg_dist",variable:="Beta"]
cgamma[variable=="delta1_v_WH1_Cg_dist",variable:="Delta"]
cgamma[variable=="delta2_v_WH1_Cg_dist",rep:=2]
cgamma[variable=="delta2_v_WH1_Cg_dist",variable:="Delta"]

#define order for plotting of bgs
cgamma$variable <- factor(cgamma$variable,levels=c("E484K","N501Y","Beta","Delta"))

ggplot(data=cgamma[rep==1 & site %in% seq(334,515),], aes(x=site, y=value, color=variable))+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("C-gamma distance versus Wuhan-Hu-1 [6m0j] (A)")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/Cgamma_v_WH1.pdf",sep=""),useDingbats=F))
```

```{r line_plots_average-atom-dist_v_WH1, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
all_atom <- melt(RBD_sites[,.(site,N501Y_v_WH1_all_atom_dist,beta1_v_WH1_all_atom_dist,beta2_v_WH1_all_atom_dist,beta3_v_WH1_all_atom_dist,delta1_v_WH1_all_atom_dist,delta2_v_WH1_all_atom_dist)], id.vars=c("site"))
all_atom[,rep:=1]
all_atom[variable=="N501Y_v_WH1_all_atom_dist",variable:="N501Y"]
all_atom[variable=="beta1_v_WH1_all_atom_dist",variable:="Beta"]
all_atom[variable=="beta2_v_WH1_all_atom_dist",rep:=2]
all_atom[variable=="beta2_v_WH1_all_atom_dist",variable:="Beta"]
all_atom[variable=="beta3_v_WH1_all_atom_dist",rep:=3]
all_atom[variable=="beta3_v_WH1_all_atom_dist",variable:="Beta"]
all_atom[variable=="delta1_v_WH1_all_atom_dist",variable:="Delta"]
all_atom[variable=="delta2_v_WH1_all_atom_dist",rep:=2]
all_atom[variable=="delta2_v_WH1_all_atom_dist",variable:="Delta"]

#define order for plotting of bgs
all_atom$variable <- factor(all_atom$variable,levels=c("E484K","N501Y","Beta","Delta"))

ggplot(data=all_atom[rep==1 & site %in% seq(334,515),], aes(x=site, y=value, color=variable))+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("Average atomic displacement versus Wuhan-Hu-1 [6m0j] (A)")+
  geom_text_repel(aes(label=ifelse(((value > 1) & variable=="Beta"),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/all_atom_v_WH1.pdf",sep=""),useDingbats=F))
```


## Compare structural shift to functional shift

We have a metric of functional perturbation at a site, derived from our deep mutational scanning measurements. Let's see how structural perturbations correlate (or don't) with structural perturbations.

```{r load_JSD}
JSD_bind <- data.table(read.csv(file=config$JSD_v_WH1_file, stringsAsFactors=F))
JSD_expr <- data.table(read.csv(file=config$JSD_v_WH1_expr_file, stringsAsFactors=F))

for(i in 1:nrow(calpha)){
  calpha$JSD_bind[i] <- JSD_bind[site==calpha$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==calpha$variable[i],JSD_min3bc]
  calpha$JSD_expr[i] <- JSD_expr[site==calpha$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==calpha$variable[i],JSD_min3bc]
}

for(i in 1:nrow(cbeta)){
  cbeta$JSD_bind[i] <- JSD_bind[site==cbeta$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==cbeta$variable[i],JSD_min3bc]
  cbeta$JSD_expr[i] <- JSD_expr[site==cbeta$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==cbeta$variable[i],JSD_min3bc]
}

for(i in 1:nrow(cgamma)){
  cgamma$JSD_bind[i] <- JSD_bind[site==cgamma$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==cgamma$variable[i],JSD_min3bc]
  cgamma$JSD_expr[i] <- JSD_expr[site==cgamma$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==cgamma$variable[i],JSD_min3bc]
}

for(i in 1:nrow(all_atom)){
  all_atom$JSD_bind[i] <- JSD_bind[site==all_atom$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==all_atom$variable[i],JSD_min3bc]
  all_atom$JSD_expr[i] <- JSD_expr[site==all_atom$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==all_atom$variable[i],JSD_min3bc]
}

```

Calpha displacement:

```{r scatter_plots_calpha_v_JSD_v_WH1, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(data=calpha[rep==1 & site %in% seq(334,515),],aes(x=abs(value),y=JSD_bind,color=variable))+
  geom_point(pch=19)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  xlab("C-alpha displacement versus Wuhan-Hu-1 [6m0j]")+
  ylab("JS distance versus Wuhan-Hu-1")+
  facet_wrap(~variable,nrow=1)+
  theme(strip.text.x = element_text(size = 18))+
  geom_text_repel(aes(label=ifelse(((JSD_bind > 0.1 & abs(value) > 0.75) | (JSD_bind > 0.15 & abs(value) > 0.25)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/scatter_JSD_v_calpha.pdf",sep=""),useDingbats=F))
```


Cbeta displacement:

```{r scatter_plots_cbeta_v_JSD_v_WH1, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(data=cbeta[rep==1 & site %in% seq(334,515),],aes(x=abs(value),y=JSD_bind,color=variable))+
  geom_point(pch=19)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  xlab("C-beta displacement versus Wuhan-Hu-1 [6m0j]")+
  ylab("JS distance versus Wuhan-Hu-1")+
  facet_wrap(~variable,nrow=1)+
  theme(strip.text.x = element_text(size = 18))+
  geom_text_repel(aes(label=ifelse(((JSD_bind > 0.1 & abs(value) > 0.75) | (JSD_bind > 0.15 & abs(value) > 0.25)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/scatter_JSD_v_cbeta.pdf",sep=""),useDingbats=F))
```

C-gamma displacement:

```{r scatter_plots_cgamma_v_JSD_v_WH1, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(data=cgamma[rep==1 & site %in% seq(334,515),],aes(x=abs(value),y=JSD_bind,color=variable))+
  geom_point(pch=19)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  xlab("C-gamma displacement versus Wuhan-Hu-1 [6m0j]")+
  ylab("JS distance versus Wuhan-Hu-1")+
  facet_wrap(~variable,nrow=1)+
  theme(strip.text.x = element_text(size = 18))+
  geom_text_repel(aes(label=ifelse(((JSD_bind > 0.1 & abs(value) > 0.75) | (JSD_bind > 0.15 & abs(value) > 0.25)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/scatter_JSD_v_cgamma.pdf",sep=""),useDingbats=F))
```

All-atom displacement:

```{r scatter_plots_all-atom_v_JSD_v_WH1, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(data=all_atom[rep==1 & site %in% seq(334,515),],aes(x=abs(value),y=JSD_bind,color=variable))+
  geom_point(pch=19)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  xlab("Average atomic displacement versus Wuhan-Hu-1 [6m0j]")+
  ylab("JS distance versus Wuhan-Hu-1")+
  facet_wrap(~variable,nrow=1)+
  theme(strip.text.x = element_text(size = 18))+
  geom_text_repel(aes(label=ifelse(((JSD_bind > 0.1 & abs(value) > 0.75) | (JSD_bind > 0.15 & abs(value) > 0.25)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/scatter_JSD_v_all_atom.pdf",sep=""),useDingbats=F))
```


# Repeat with aligned cryo-EM based local refined RBD:ACE2 structures

## Setup

Read in PDBs. These were aligned to minimize RBD Calpha RMSD using PyMol. Structures include WH1 (7kmb), N501Y (7ekf), beta (7vx4), and delta (7v8b)

```{r input_data_cryo}
WH1 <- read.pdb(file=config$pdb_WH1_cryo)
alpha <- read.pdb(file=config$pdb_alpha_cryo)
beta <- read.pdb(file=config$pdb_beta_cryo)
delta <- read.pdb(file=config$pdb_delta_cryo)
```

## Calculate pairwise C-alpha, beta, and gamma distances for each structure compared to WH1

Note, the Cgamma calc excludes residues with no gamma-atom (alanine and glycine) or beta-branching (and thus a gamma1 and gamma2 atom: isoleucine, threonine, valine)

```{r calc_Ca_dists_cryo}
RBD_sites <- data.table(read.csv(file=config$RBD_sites))

for(voc in c("alpha","beta","delta")){
  RBD_sites[,paste(voc,"_v_WH1_Ca_dist",sep=""):=as.numeric(NA)]
  for(i in 1:nrow(RBD_sites)){
    site <- RBD_sites[i,site]
    WH1_calpha <- WH1$atom[WH1$atom$resno==site & WH1$atom$elety=="CA",]
    voc_calpha <- get(voc)$atom[get(voc)$atom$resno==site & get(voc)$atom$elety=="CA",]
    if(nrow(WH1_calpha)==1 & nrow(voc_calpha)==1){
      RBD_sites[i,paste(voc,"_v_WH1_Ca_dist",sep="") := sqrt((WH1_calpha$x - voc_calpha$x)^2 + (WH1_calpha$y - voc_calpha$y)^2 + (WH1_calpha$z - voc_calpha$z)^2)] 
    }else{
      #print(paste("no distance for", voc, "site",site))
    }
  }
}

```

```{r calc_Cb_dists_cryo}
for(voc in c("alpha","beta","delta")){
  RBD_sites[,paste(voc,"_v_WH1_Cb_dist",sep=""):=as.numeric(NA)]
  for(i in 1:nrow(RBD_sites)){
    site <- RBD_sites[i,site]
    WH1_cbeta <- WH1$atom[WH1$atom$resno==site & WH1$atom$elety=="CB",]
    voc_cbeta <- get(voc)$atom[get(voc)$atom$resno==site & get(voc)$atom$elety=="CB",]
    if(nrow(WH1_cbeta)==1 & nrow(voc_cbeta)==1){
      RBD_sites[i,paste(voc,"_v_WH1_Cb_dist",sep="") := sqrt((WH1_cbeta$x - voc_cbeta$x)^2 + (WH1_cbeta$y - voc_cbeta$y)^2 + (WH1_cbeta$z - voc_cbeta$z)^2)] 
    }else{
      #print(paste("no distance for", voc, "site",site))
    }
  }
}

```

```{r calc_Cg_dists_cryo}
for(voc in c("alpha","beta","delta")){
  RBD_sites[,paste(voc,"_v_WH1_Cg_dist",sep=""):=as.numeric(NA)]
  for(i in 1:nrow(RBD_sites)){
    site <- RBD_sites[i,site]
    WH1_cgamma <- WH1$atom[WH1$atom$resno==site & WH1$atom$elety %in% c("CG","SG","OG"),]
    voc_cgamma <- get(voc)$atom[get(voc)$atom$resno==site & get(voc)$atom$elety %in% c("CG","SG","OG"),]
    if(nrow(WH1_cgamma)==1 & nrow(voc_cgamma)==1){
      RBD_sites[i,paste(voc,"_v_WH1_Cg_dist",sep="") := sqrt((WH1_cgamma$x - voc_cgamma$x)^2 + (WH1_cgamma$y - voc_cgamma$y)^2 + (WH1_cgamma$z - voc_cgamma$z)^2)] 
    }else{
      #print(paste("no distance for", voc, "site",site))
    }
  }
}

```

Average across all atoms in each residue

```{r calc_all-atom_dists_cryo}
for(voc in c("alpha","beta","delta")){
  RBD_sites[,paste(voc,"_v_WH1_all_atom_dist",sep=""):=as.numeric(NA)]
  for(i in 1:nrow(RBD_sites)){
    site <- RBD_sites[i,site]
    WH1_atoms <- WH1$atom[WH1$atom$resno==site,]
    voc_atoms <- get(voc)$atom[get(voc)$atom$resno==site,]
    if(nrow(WH1_atoms)>1 & nrow(voc_atoms)>1 & nrow(WH1_atoms)==nrow(voc_atoms)){
      if(WH1_atoms$resid[1]==voc_atoms$resid[1]){
        RBD_sites[i,paste(voc,"_v_WH1_all_atom_dist",sep="") := mean(sqrt((WH1_atoms$x - voc_atoms$x)^2 + (WH1_atoms$y - voc_atoms$y)^2 + (WH1_atoms$z - voc_atoms$z)^2))] 
      }
    }
  }
}

```

Line plots of Calpha, beta, gamma displacement compared to WH1 structure

```{r line_plots_Calpha-dist_v_WH1_cryo, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
calpha <- melt(RBD_sites[,.(site,alpha_v_WH1_Ca_dist,beta_v_WH1_Ca_dist,delta_v_WH1_Ca_dist)], id.vars=c("site"))
calpha[variable=="alpha_v_WH1_Ca_dist",variable:="N501Y"]
calpha[variable=="beta_v_WH1_Ca_dist",variable:="Beta"]
calpha[variable=="delta_v_WH1_Ca_dist",variable:="Delta"]

#define colors for each bg
group.colors <- c("Wuhan-Hu-1" = cbPalette[1], "N501Y" = cbPalette[3], "E484K" = cbPalette[5], "Beta"=cbPalette[6], "Delta"=cbPalette[7], "Omicron"=cbPalette[2])

#define order for plotting of bgs
calpha$variable <- factor(calpha$variable,levels=c("E484K","N501Y","Beta","Delta"))

ggplot(data=calpha[site %in% seq(334,515),], aes(x=site, y=value, color=variable))+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("C-alpha distance versus Wuhan-Hu-1 [7kmb] (A)")+
  geom_text_repel(aes(label=ifelse(((value > 1) & variable=="Beta"),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/Calpha_v_WH1_cryo.pdf",sep=""),useDingbats=F))
```

```{r line_plots_Cbeta-dist_v_WH1_cryo, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
cbeta <- melt(RBD_sites[,.(site,alpha_v_WH1_Cb_dist,beta_v_WH1_Cb_dist,delta_v_WH1_Cb_dist)], id.vars=c("site"))
cbeta[variable=="alpha_v_WH1_Cb_dist",variable:="N501Y"]
cbeta[variable=="beta_v_WH1_Cb_dist",variable:="Beta"]
cbeta[variable=="delta_v_WH1_Cb_dist",variable:="Delta"]

#define order for plotting of bgs
cbeta$variable <- factor(cbeta$variable,levels=c("E484K","N501Y","Beta","Delta"))

ggplot(data=cbeta[site %in% seq(334,515),], aes(x=site, y=value, color=variable))+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("C-beta distance versus Wuhan-Hu-1 [7kmb] (A)")+
  geom_text_repel(aes(label=ifelse(((value > 2) & variable=="Beta"),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/Cbeta_v_WH1_cryo.pdf",sep=""),useDingbats=F))
```


```{r line_plots_Cgamma-dist_v_WH1_cryo, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
cgamma <- melt(RBD_sites[,.(site,alpha_v_WH1_Cg_dist,beta_v_WH1_Cg_dist,delta_v_WH1_Cg_dist)], id.vars=c("site"))
cgamma[variable=="alpha_v_WH1_Cg_dist",variable:="N501Y"]
cgamma[variable=="beta_v_WH1_Cg_dist",variable:="Beta"]
cgamma[variable=="delta_v_WH1_Cg_dist",variable:="Delta"]

#define order for plotting of bgs
cgamma$variable <- factor(cgamma$variable,levels=c("E484K","N501Y","Beta","Delta"))

ggplot(data=cgamma[site %in% seq(334,515),], aes(x=site, y=value, color=variable))+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("C-gamma distance versus Wuhan-Hu-1 [7kmb] (A)")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/Cgamma_v_WH1_cryo.pdf",sep=""),useDingbats=F))
```

```{r line_plots_all_atom-dist_v_WH1_cryo, echo=T, fig.width=12, fig.height=4, fig.align="center", dpi=300,dev="png"}
#define focal bg for others to compare to
all_atom <- melt(RBD_sites[,.(site,alpha_v_WH1_all_atom_dist,beta_v_WH1_all_atom_dist,delta_v_WH1_all_atom_dist)], id.vars=c("site"))
all_atom[variable=="alpha_v_WH1_all_atom_dist",variable:="N501Y"]
all_atom[variable=="beta_v_WH1_all_atom_dist",variable:="Beta"]
all_atom[variable=="delta_v_WH1_all_atom_dist",variable:="Delta"]

#define order for plotting of bgs
all_atom$variable <- factor(all_atom$variable,levels=c("E484K","N501Y","Beta","Delta"))

ggplot(data=all_atom[site %in% seq(334,515),], aes(x=site, y=value, color=variable))+
  geom_line(size=1)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  scale_x_continuous(expand=c(0.01,0.01),breaks=c(331,seq(335,530,by=5)))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.6,face="bold",size=10))+
  ylab("Average atomic displacement versus Wuhan-Hu-1 [7kmb] (A)")+
  geom_text_repel(aes(label=ifelse(((value > 2) & variable=="Beta"),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/all_atom_v_WH1_cryo.pdf",sep=""),useDingbats=F))
```

## Compare structural shift to functional shift

We have a metric of functional perturbation at a site, derived from our deep mutational scanning measurements. Let's see how structural perturbations correlate (or don't) with structural perturbations.

```{r load_JSD_cryo}
JSD_bind <- data.table(read.csv(file=config$JSD_v_WH1_file, stringsAsFactors=F))
JSD_expr <- data.table(read.csv(file=config$JSD_v_WH1_expr_file, stringsAsFactors=F))

for(i in 1:nrow(calpha)){
  calpha$JSD_bind[i] <- JSD_bind[site==calpha$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==calpha$variable[i],JSD_min3bc]
  calpha$JSD_expr[i] <- JSD_expr[site==calpha$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==calpha$variable[i],JSD_min3bc]
}

for(i in 1:nrow(cbeta)){
  cbeta$JSD_bind[i] <- JSD_bind[site==cbeta$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==cbeta$variable[i],JSD_min3bc]
  cbeta$JSD_expr[i] <- JSD_expr[site==cbeta$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==cbeta$variable[i],JSD_min3bc]
}

for(i in 1:nrow(cgamma)){
  cgamma$JSD_bind[i] <- JSD_bind[site==cgamma$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==cgamma$variable[i],JSD_min3bc]
  cgamma$JSD_expr[i] <- JSD_expr[site==cgamma$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==cgamma$variable[i],JSD_min3bc]
}

for(i in 1:nrow(all_atom)){
  all_atom$JSD_bind[i] <- JSD_bind[site==all_atom$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==all_atom$variable[i],JSD_min3bc]
  all_atom$JSD_expr[i] <- JSD_expr[site==all_atom$site[i] & bg_1=="Wuhan-Hu-1" & bg_2==all_atom$variable[i],JSD_min3bc]
}

```

Calpha displacement:

```{r scatter_plots_calpha_v_JSD_v_WH1_cryo, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(data=calpha[site %in% seq(334,515),],aes(x=abs(value),y=JSD_bind,color=variable))+
  geom_point(pch=19)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  xlab("C-alpha displacement versus Wuhan-Hu-1 [7kmb]")+
  ylab("JS distance versus Wuhan-Hu-1")+
  facet_wrap(~variable,nrow=1)+
  theme(strip.text.x = element_text(size = 18))+
  geom_text_repel(aes(label=ifelse(((JSD_bind > 0.1 & abs(value) > 0.75) | (JSD_bind > 0.15 & abs(value) > 0.25)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/scatter_JSD_v_calpha_cryo.pdf",sep=""),useDingbats=F))
```


Cbeta displacement:

```{r scatter_plots_cbeta_v_JSD_v_WH1_cryo, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(data=cbeta[site %in% seq(334,515),],aes(x=abs(value),y=JSD_bind,color=variable))+
  geom_point(pch=19)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  xlab("C-beta displacement versus Wuhan-Hu-1 [7kmb]")+
  ylab("JS distance versus Wuhan-Hu-1")+
  facet_wrap(~variable,nrow=1)+
  theme(strip.text.x = element_text(size = 18))+
  geom_text_repel(aes(label=ifelse(((JSD_bind > 0.1 & abs(value) > 0.75) | (JSD_bind > 0.15 & abs(value) > 0.25)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/scatter_JSD_v_cbeta_cryo.pdf",sep=""),useDingbats=F))
```

C-gamma displacement:

```{r scatter_plots_cgamma_v_JSD_v_WH1_cryo, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(data=cgamma[site %in% seq(334,515),],aes(x=abs(value),y=JSD_bind,color=variable))+
  geom_point(pch=19)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  xlab("C-gamma displacement versus Wuhan-Hu-1 [7kmb]")+
  ylab("JS distance versus Wuhan-Hu-1")+
  facet_wrap(~variable,nrow=1)+
  theme(strip.text.x = element_text(size = 18))+
  geom_text_repel(aes(label=ifelse(((JSD_bind > 0.1 & abs(value) > 0.75) | (JSD_bind > 0.15 & abs(value) > 0.25)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/scatter_JSD_v_cgamma_cryo.pdf",sep=""),useDingbats=F))
```

```{r scatter_plots_all_atom_v_JSD_v_WH1_cryo, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(data=all_atom[site %in% seq(334,515),],aes(x=abs(value),y=JSD_bind,color=variable))+
  geom_point(pch=19)+
  scale_color_manual(values=group.colors)+
  theme_classic()+
  xlab("Average atomic displacement versus Wuhan-Hu-1 [7kmb]")+
  ylab("JS distance versus Wuhan-Hu-1")+
  facet_wrap(~variable,nrow=1)+
  theme(strip.text.x = element_text(size = 18))+
  geom_text_repel(aes(label=ifelse(((JSD_bind > 0.1 & abs(value) > 0.75) | (JSD_bind > 0.15 & abs(value) > 0.25)),as.character(site),'')),size=3,color="gray40")

invisible(dev.print(pdf, paste(config$structural_shifts_dir,"/scatter_JSD_v_all_atom_cryo.pdf",sep=""),useDingbats=F))
```
