---
title: "Titer Spike pseudovirus by flow cytometry"
output:
  github_document:
    html_preview: no
editor_options:
  chunk_output_type: inline
---

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
#list of packages to install/load
packages = c("ggplot2", "data.table", "tidyverse", "dplyr", "gridExtra")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))
#make results directory
if(!file.exists("results")){
 dir.create(file.path("results"))
}

```

## Experiment: Titer Spike point mutant pseudovirus

**2022-01-17 (ACE2hi)**
**2022-01-24 (ACE2lo)**
**2022-01-17 (p24 ELISA)**
**2022-02-18 (ACE2lo [reps 4-6]), by ANL and RE**

### Define colorblind-friendly palette
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", 
               "#0072B2", "#D55E00", "#CC79A7")
# The palette with black
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                "#0072B2", "#D55E00", "#CC79A7")
```

### Read in data from titering (luciferase) and p24 quantifications
```{r read_input}
dt <- read.csv(file="./titer_data/all_data.csv", stringsAsFactors=F)
head(dt)
```

### Plot absolute (RLU/uL) titers as measured by luciferase activity for WH1 cycle, HEK-ACE2 cell line
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
dt_WH1 <- dt[dt$background=="WH1_D614G",] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "Y449H", "N501Y", "Y449H/N501Y")))

ggplot(dt_WH1,
       aes(genotype, RLU_per_uL_ACE2hi)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(1e0, 1e6)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/uL)") +
  theme_bw()  +
  ggtitle("D614G background, ACE2-high cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+
  stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_D614G-cycle_absolute_ACE2hi.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)
```


### Plot absolute (RLU/uL) titers as measured by luciferase activity for omicron cycle, HEK-ACE2 cell line
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
dt_omicron_rep1 <- dt[dt$background=="omicron" & dt$rep %in% c(1,2,3),] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "R498Q", "Y501N", "R498Q/Y501N")))
dt_omicron_rep2 <- dt[dt$background=="omicron" & dt$rep %in% c(4,5,6),] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "R498Q", "Y501N", "R498Q/Y501N")))
dt_omicron_all <- dt[dt$background=="omicron",] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "R498Q", "Y501N", "R498Q/Y501N")))

ggplot(dt_omicron_rep1,
       aes(genotype, RLU_per_uL_ACE2hi)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(4e-2, 1e6)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/uL)") +
  theme_bw()  +
  ggtitle("Omicron background, ACE2-high cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+
  stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers-omicron-cycle_absolute_ACE2hi.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)
```

### Plot absolute (RLU/uL) titers as measured by luciferase activity for WH1 cycle, "clone A" cell line
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(dt_WH1,
       aes(genotype, RLU_per_uL_ACE2lo)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(1e0, 1e5)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/uL)") +
  theme_bw()  +
  ggtitle("D614G background, ACE2-low cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_D614G-cycle_absolute_ACE2lo.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)
```


### Plot absolute (RLU/uL) titers as measured by luciferase activity for omicron cycle, "clone A" cell line, two reps and pooled
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(dt_omicron_rep1,
       aes(genotype, RLU_per_uL_ACE2lo)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(1e0, 5e4)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/uL)") +
  theme_bw()  +
  ggtitle("Omicron background, ACE2-low cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_omicron-cycle1_absolute_ACE2lo.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)
```
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(dt_omicron_rep2,
       aes(genotype, RLU_per_uL_ACE2lo)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(5e-1, 5e4)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/uL)") +
  theme_bw()  +
  ggtitle("Omicron background, ACE2-low cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_omicron-cycle2_absolute_ACE2lo.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)
```

### Plot titers as RLU per pg p24 for WH1 cycle, HEK-ACE2 cell line

```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
dt$RLU_per_pg_p24_ACE2hi <- dt$RLU_per_uL_ACE2hi * 1000 / dt$p24_pg_per_mL
dt$RLU_per_pg_p24_ACE2lo <- dt$RLU_per_uL_ACE2lo * 1000 / dt$p24_pg_per_mL
```

```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
dt_WH1 <- dt[dt$background=="WH1_D614G",] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "Y449H", "N501Y", "Y449H/N501Y")))

ggplot(dt_WH1,
       aes(genotype, RLU_per_pg_p24_ACE2hi)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(1e-3, 1e3)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/pg p24)") +
  theme_bw()  +
  ggtitle("D614G background, ACE2-high cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_D614G-cycle_RLU-per-pg-p24_ACE2hi.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)

```


### Plot titers as RLU per pg p24 for WH1 cycle, "clone A" cell line
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
dt_WH1 <- dt[dt$background=="WH1_D614G",] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "Y449H", "N501Y", "Y449H/N501Y")))

ggplot(dt_WH1,
       aes(genotype, RLU_per_pg_p24_ACE2lo)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(1e-3, 1e2)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/pg p24)") +
  theme_bw()  +
  ggtitle("D614G background, ACE2-low cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_D614G-cycle_RLU-per-pg-p24_ACE2lo.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)

```

### Plot titers as RLU per pg p24 for Omicron cycle, HEK-ACE2 cell line

```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
dt_omicron_rep1 <- dt[dt$background=="omicron" & dt$rep %in% c(1,2,3),] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "R498Q", "Y501N", "R498Q/Y501N")))
dt_omicron_rep2 <- dt[dt$background=="omicron" & dt$rep %in% c(4,5,6),] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "R498Q", "Y501N", "R498Q/Y501N")))
dt_omicron_all <- dt[dt$background=="omicron",] %>% mutate(genotype = factor(genotype, 
                                levels=c("none", "wt", 
                                         "R498Q", "Y501N", "R498Q/Y501N")))

ggplot(dt_omicron_rep1,
       aes(genotype, RLU_per_pg_p24_ACE2hi)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(1e-5, 1e3)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/pg p24)") +
  theme_bw()  +
  ggtitle("Omicron background, ACE2-high cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_omicron-cycle_RLU-per-pg-p24_ACE2hi.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)

```


### Plot titers as RLU per pg p24 for omicron cycle, "clone A" cell line
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(dt_omicron_rep1,
       aes(genotype, RLU_per_pg_p24_ACE2lo)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(8e-4, 1e2)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/pg p24)") +
  theme_bw()  +
  ggtitle("Omicron background, ACE2-low cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_omicron-cycle1_RLU-per-pg-p24_ACE2lo.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)

```
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(dt_omicron_rep2,
       aes(genotype, RLU_per_pg_p24_ACE2lo)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(1e-4, 1e2)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/pg p24)") +
  theme_bw()  +
  ggtitle("Omicron background, ACE2-low cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_omicron-cycle2_RLU-per-pg-p24_ACE2lo.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)

```
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}
ggplot(dt_omicron_all,
       aes(genotype, RLU_per_pg_p24_ACE2lo)) +
  geom_point(size=3.5,
             shape=16,
             position=position_jitter(width = 0.1, height = 0, seed = 0)) +
  #geom_hline(yintercept=25/exp(mean(log(dt[dt$genotype=="wildtype" & dt$uL_virus=="50","virus_per_mL_stringent"]))), color="grey", linetype="dashed") + #Limit of detection
  scale_y_log10(lim=c(1e-4, 1e2)) +
  xlab("Spike gene") +
  ylab("entry titer (RLU/pg p24)") +
  theme_bw()  +
  ggtitle("Omicron background, ACE2-low cell line") +
  theme(plot.title = element_text(hjust = 0.5),text=element_text(size=10))+   stat_summary(fun=mean,geom="point",size=10,color="red",shape=95)

ggsave(
  "./results/titers_omicron-cycleall_RLU-per-pg-p24_ACE2lo.pdf",
  scale = 1,
  width = 3.5,
  height = 3,
  useDingbats=F
)

```
### quants and t-tests for titer differences in omicron cycle
```{r, fig.width=3.5, fig.height=3, fig.align="center", dpi=300,dev="png"}


#calc fold-change in geomean titer compared to wt, absolute titer, ACE2hi
#omicron R498Q vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_uL_ACE2hi"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2hi"])
#omicron Y501N vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_uL_ACE2hi"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2hi"])
#omicron R498Q/Y501N vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_uL_ACE2hi"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2hi"])

#calc fold-change in geomean titer compared to wt, absolute titer, ACE2lo
#omicron R498Q vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_uL_ACE2lo"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2lo"])
#omicron Y501N vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_uL_ACE2lo"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2lo"])
#omicron R498Q/Y501N vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_uL_ACE2lo"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2lo"])

#calc fold-change in geomean titer compared to wt, RLU/pg p24 titer, ACE2hi
#omicron R498Q vs wt
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_pg_p24_ACE2hi"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2hi"])
#omicron Y501N vs wt
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_pg_p24_ACE2hi"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2hi"])
#omicron R498Q/Y501N vs wt
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2hi"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2hi"])

#calc fold-change in geomean titer compared to wt, RLU/pg p24 titer, ACE2lo
#omicron R498Q vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2lo"])
#omicron Y501N vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2lo"])
#omicron R498Q/Y501N vs wt, rep1
geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2lo"])

#calc fold-change in geomean titer compared to wt, RLU/pg p24 titer, ACE2lo
#omicron R498Q vs wt, rep2
geomean(dt_omicron_rep2[dt_omicron_rep2$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_rep2[dt_omicron_rep2$genotype=="wt","RLU_per_pg_p24_ACE2lo"])
#omicron Y501N vs wt, rep2
geomean(dt_omicron_rep2[dt_omicron_rep2$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_rep2[dt_omicron_rep2$genotype=="wt","RLU_per_pg_p24_ACE2lo"])
#omicron R498Q/Y501N vs wt, rep2
geomean(dt_omicron_rep2[dt_omicron_rep2$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_rep2[dt_omicron_rep2$genotype=="wt","RLU_per_pg_p24_ACE2lo"])

#calc fold-change in geomean titer compared to wt, RLU/pg p24 titer, ACE2lo
#omicron R498Q vs wt, all reps
geomean(dt_omicron_all[dt_omicron_all$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_all[dt_omicron_all$genotype=="wt","RLU_per_pg_p24_ACE2lo"])
#omicron Y501N vs wt, all reps
geomean(dt_omicron_all[dt_omicron_all$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_all[dt_omicron_all$genotype=="wt","RLU_per_pg_p24_ACE2lo"])
#omicron R498Q/Y501N vs wt, all reps
geomean(dt_omicron_all[dt_omicron_all$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"])/geomean(dt_omicron_all[dt_omicron_all$genotype=="wt","RLU_per_pg_p24_ACE2lo"])

#calc fold-change in geomean titer compared to wt, absolute titer, ACE2hi
#WH1 Y449H vs wt
geomean(dt_WH1[dt_WH1$genotype=="Y449H","RLU_per_uL_ACE2hi"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_uL_ACE2hi"])
#WH1 N501Y vs wt
geomean(dt_WH1[dt_WH1$genotype=="N501Y","RLU_per_uL_ACE2hi"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_uL_ACE2hi"])
#WH1 Y449H/N501Y vs wt
geomean(dt_WH1[dt_WH1$genotype=="Y449H/N501Y","RLU_per_uL_ACE2hi"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_uL_ACE2hi"])

#calc fold-change in geomean titer compared to wt, absolute titer, ACE2lo
#WH1 Y449H vs wt
geomean(dt_WH1[dt_WH1$genotype=="Y449H","RLU_per_uL_ACE2lo"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_uL_ACE2lo"])
#WH1 N501Y vs wt
geomean(dt_WH1[dt_WH1$genotype=="N501Y","RLU_per_uL_ACE2lo"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_uL_ACE2lo"])
#WH1 Y449H/N501Y vs wt
geomean(dt_WH1[dt_WH1$genotype=="Y449H/N501Y","RLU_per_uL_ACE2lo"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_uL_ACE2lo"])

#calc fold-change in geomean titer compared to wt, RLU/pg p24 titer, ACE2hi
#WH1 Y449H vs wt
geomean(dt_WH1[dt_WH1$genotype=="Y449H","RLU_per_pg_p24_ACE2hi"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_pg_p24_ACE2hi"])
#WH1 N501Y vs wt
geomean(dt_WH1[dt_WH1$genotype=="N501Y","RLU_per_pg_p24_ACE2hi"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_pg_p24_ACE2hi"])
#WH1 Y449H/N501Y vs wt
geomean(dt_WH1[dt_WH1$genotype=="Y449H/N501Y","RLU_per_pg_p24_ACE2hi"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_pg_p24_ACE2hi"])

#calc fold-change in geomean titer compared to wt, RLU/pg p24 titer, ACE2lo
#WH1 Y449H vs wt
geomean(dt_WH1[dt_WH1$genotype=="Y449H","RLU_per_pg_p24_ACE2lo"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_pg_p24_ACE2lo"])
#WH1 N501Y vs wt
geomean(dt_WH1[dt_WH1$genotype=="N501Y","RLU_per_pg_p24_ACE2lo"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_pg_p24_ACE2lo"])
#WH1 Y449H/N501Y vs wt
geomean(dt_WH1[dt_WH1$genotype=="Y449H/N501Y","RLU_per_pg_p24_ACE2lo"])/geomean(dt_WH1[dt_WH1$genotype=="wt","RLU_per_pg_p24_ACE2lo"])



t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_pg_p24_ACE2hi"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_pg_p24_ACE2hi"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2hi"]))

t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_pg_p24_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_pg_p24_ACE2hi"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_pg_p24_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2hi"]))

t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_pg_p24_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2hi"]))

#rep1, ACE2lo, omicron cycle t tests
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))

t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))

t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))

#rep2, ACE2lo, omicron cycle t tests
t.test(log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))

t.test(log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))

t.test(log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_rep2[dt_omicron_rep2$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))

#all reps, ACE2lo, omicron cycle t tests
t.test(log10(dt_omicron_all[dt_omicron_all$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_all[dt_omicron_all$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_all[dt_omicron_all$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_all[dt_omicron_all$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_all[dt_omicron_all$genotype=="wt","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_all[dt_omicron_all$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))

t.test(log10(dt_omicron_all[dt_omicron_all$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_all[dt_omicron_all$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]))
t.test(log10(dt_omicron_all[dt_omicron_all$genotype=="R498Q","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_all[dt_omicron_all$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))

t.test(log10(dt_omicron_all[dt_omicron_all$genotype=="Y501N","RLU_per_pg_p24_ACE2lo"]),log10(dt_omicron_all[dt_omicron_all$genotype=="R498Q/Y501N","RLU_per_pg_p24_ACE2lo"]))


t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_uL_ACE2hi"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_uL_ACE2hi"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_uL_ACE2hi"]))

t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_uL_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_uL_ACE2hi"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_uL_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_uL_ACE2hi"]))

t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_uL_ACE2hi"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_uL_ACE2hi"]))


t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_uL_ACE2lo"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_uL_ACE2lo"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="wt","RLU_per_uL_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_uL_ACE2lo"]))

t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_uL_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_uL_ACE2lo"]))
t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q","RLU_per_uL_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_uL_ACE2lo"]))

t.test(log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="Y501N","RLU_per_uL_ACE2lo"]),log10(dt_omicron_rep1[dt_omicron_rep1$genotype=="R498Q/Y501N","RLU_per_uL_ACE2lo"]))


```